---
title: 'Deaths in the United State Report'
author: 'Angelica Gamboa and Gwendolyn Espinoza'
output: html_document
date: "2025-04-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# INTRODUCTION

- The data was gathered from **National Center for Health Statistics (NCHS)** – *Leading Causes of Death in the U.S.* This data gives us the top 10 most causes of deaths in the United States. We are given the data from every state in the United States and the timing is from 1999 to 2017. Through this data we are able to see the progression of the United States over time due to the amount of deaths in each year. 


# QUESTIONS
1. What are the leading causes of death since 2001?
2. Which state had the most amount of deaths in the US?
3. What was the leading cause of death in California?
4. How has the cause of heart disease deaths changed over time?
5. How has the cause of cancer deaths changed over time?
6. We know the adjusted death rate for the data, but what does the rate look like when we subtract "all causes'?
7. What is the death rate by cause and state?
8. What is the age-adjusted death rate for heart disease?


```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
library(usmap)
library(scales)
data <- read.csv("NCHS_-_Leading_Causes_of_Death__United_States (1).csv")
```


# Variables in the dataset:
  - Year: `Year in which the incident took place.`
  - X113 cause name: `The full, official name of the cause of death classified by the NCHS.`
  - Cause name: `A broader label for the cause of death.`
  - State: `The U.S. state where the data was collected.`
  - Deaths: `The total number of deaths reported for a specific cause of death.`
  - Age-adjusted death rate: `The death rate per 100,000 people, adjusted per age group for fair comparison.`


## **Leading Causes of Death Since 2001**

```{r echo = FALSE}
# Filtering the data and removing "United States" from the column named 'States'
top_causes <- data %>%
  filter(Cause.Name != "All causes", State != "United States") %>% 
  group_by(Cause.Name) %>%
  summarise(total_deaths = sum(Deaths)) %>%
  arrange(desc(total_deaths))
  
# Plotting causes of death
ggplot(top_causes, aes(x = total_deaths, y = reorder(Cause.Name, total_deaths))) +
  geom_col(fill = 'purple3') +
  labs(title = 'Top 10 Causes of Death in the U.S. Since 1999',
       x = 'Total Deaths',
       y = 'Cause of Death',
       subtitle = 'CLRD stands for Chronic Lower Respiratory Diseases')
```
This is the total summation of all Deaths from 1999 to 2017. We can see that there are many deaths due to Heart Disease and Cancer. The least amount of deaths being by suicide.  


## **Total Deaths in Each State**

```{r echo = FALSE}
library(dplyr)
library(ggplot2)

death_totals <- data %>% 
  select(State, Deaths, Cause.Name) %>%
  filter(State != 'United States', Cause.Name != 'All causes') %>%
  group_by(State) %>%
  summarize(Total_Deaths = sum(Deaths, na.rm = TRUE))


# Highlighting California since it has the most deaths
death_totals <- death_totals %>%
  mutate(Color = ifelse(State == "California", "California", "Other"))

TOTAL_DEATHS_BY_STATE <- ggplot(death_totals, aes(x = Total_Deaths, y = reorder(State, Total_Deaths), fill = Color)) +
  geom_col() +
  scale_fill_manual(values = c("California" = "red")) +
  labs(
    x = "Deaths",
    y = "States",
    title = "Total Amount of Deaths in Each State Since 2001"
  ) +
  theme_minimal() +
  theme(legend.position = "none") 

print(TOTAL_DEATHS_BY_STATE)
```

```{r echo = FALSE}

US_TOTAL_DEATHS <- data %>%
  filter(State == "United States") %>%
  select(Year, Cause.Name, US_Deaths = Deaths)

ALL_CAUSE_DEATH <- data %>%
  filter(Cause.Name == "All causes") %>%
  select(Year, State, All_Causes_Deaths = Deaths)

BOTH_COMBINED <- data %>%
  left_join(US_TOTAL_DEATHS, by = c("Year", "Cause.Name")) %>%
  left_join(ALL_CAUSE_DEATH, by = c("Year", "State")) %>%
  mutate(
    `Percentage of Deaths` = ((Deaths / All_Causes_Deaths) * 100))


PLOTTING_PERCENTAGES <- BOTH_COMBINED %>% filter(State != 'United States',Cause.Name != 'All causes') %>%
  ggplot(aes ( y = State, x = `Percentage of Deaths`, color = Cause.Name )) + geom_point() + labs(
    title = 'Death percentage based on each Cause'  )



PLOTTING_PERC <- BOTH_COMBINED %>% filter(Cause.Name != 'All causes')  %>%
  ggplot(aes(x = Cause.Name, y = State, fill = `Percentage of Deaths`)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "C") +
  labs(title = "Heatmap of Death Percentages by Cause and State", x = "Cause", y = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_discrete(expand = expansion(mult = c(.005, 0.005))) + theme(axis.text.y = element_text(size = 5))


PLOTTING_PERC

#PLOTTING_PERCENTAGES




total_deaths <- data %>%
  filter(Cause.Name == "All causes") %>%
  group_by(State) %>%
  summarise(Total_Deaths = sum(Deaths, na.rm = TRUE))

cause_deaths <- data %>%
  filter(Cause.Name != "All causes") %>%
  group_by(State) %>%
  summarize(Cause_Deaths = sum(Deaths, na.rm = TRUE))

merged <- left_join(cause_deaths, total_deaths, by = "State") %>%
  mutate(Proportion = 100 * Cause_Deaths / Total_Deaths)

merged_map <- merged %>%
  mutate(state = tolower(State)) %>%
  select(state, Proportion)

plot_usmap(data = merged_map, values = "Proportion", labels = FALSE) +
  scale_fill_continuous(name = "Death Percentages", label = percent_format(scale = 1)) +
  labs(title = "Total Cause Deaths over All Cause Deaths by State") +
  theme(legend.position = "right")
```


## **Leading Cause of Deaths in California **

```{r echo=FALSE}
Leading_cause <- data %>%
  filter(State == 'California', Cause.Name != 'All causes') %>%
  group_by(Cause.Name) %>%
  summarize(sum(Deaths)) %>%
  ggplot(aes(x = Cause.Name, y = `sum(Deaths)`)) + geom_col(fill = 'violet', color = 'black') +
  labs(
    title = "Leading Causes of Death in California",
    x = "Cause Name",
    y = "Total Number of Deaths"
  ) +
  theme_minimal() +
  scale_x_discrete(guide = guide_axis( angle = 80))

Leading_cause
```


## Changes Throughout the Years

```{r echo = FALSE}
#HIGHER PERCENTAGE = SUGGESTS AN OLDER POPULATION


#ALL DISEASES SEE HOW THEY ALL INCREASE / DECREASE THROUGHOUT THE YEARS
ALL <- data %>%
  select(Year, Cause.Name, Deaths) %>%
  filter(Cause.Name != 'All causes') %>%
  group_by(Cause.Name, Year) %>%
  summarize(TOT = sum(Deaths)) %>%
  ggplot(aes(x = Year, y = TOT, color = Cause.Name)) + geom_point(alpha = .5) + geom_line()+ labs(
   x = 'Years',
   y = 'Total Amount of Deaths',
   title = 'All Causes Throughout the Years',
   color = 'Causes'
    )

 
ALL
```


## **How has the number of heart disease deaths changed over time?**

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(kableExtra)

# Filtering the heart disease data
heart_data <- subset(data, grepl("Heart disease", Cause.Name, ignore.case = TRUE))

# Number of heart disease deaths each year since 1999
heart_trend <- heart_data %>%
  group_by(Year) %>%
  summarise(Total_Heart_Deaths = sum(Deaths, na.rm = TRUE))

heart_trend %>%
  mutate(Total_Heart_Deaths = format(Total_Heart_Deaths, big.mark = ",")) %>%
  kable(col.names = c("Year", "Total Deaths")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width = FALSE,
    position = "left"
  ) %>%
  column_spec(1, width = "4cm", bold = TRUE) %>%
  column_spec(2, width = "4cm")        

# Number of heart disease deaths each year since 2001
HEART_DISEASE <- data %>% 
  filter(Cause.Name == 'Heart disease') %>%
  group_by(Year) %>%
  summarize(TOTAL = sum(Deaths)) %>%
  
  # Plotting the rate of heart disease deaths over time
  ggplot(aes(x = Year, y = TOTAL)) + 
  geom_line(color = 'lightblue', linewidth = 1.2) + 
  geom_point(color = 'black') +
  labs(
    title = "Total Heart Disease Deaths",
    x = "Year",
    y = "Total number of Heart Disease Deaths"
  ) +
  theme_minimal()

HEART_DISEASE
```

This plot show how Heart Disease has been decreasing throughout time. The United States started bringing in more awareness to Heart Disease and people started reaching out for help. We can see that from 2010 and 2012, we see a dip in deaths, this is due to the Million Hearts being launched by the United States Department of Health. Their main goal was to prevent 1 million heart diseases by 2017. We can see this dip in the plot above. Before 2010, This can be due to people making better choices for themselves, such as not smoking and better eating habits.

## **How has the number of cancer deaths changed over time?**

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)

# Filtering the cancer data
cancer_data <- subset(data, grepl("cancer", Cause.Name, ignore.case = TRUE))

# Number of cancer deaths each year since 2001
cancer_trend <- cancer_data %>%
  group_by(Year) %>%
  summarise(Total_Deaths = sum(Deaths))

library(kableExtra)

cancer_trend %>%
  mutate(Total_Deaths = format(Total_Deaths, big.mark = ",")) %>%
  kable(col.names = c("Year", "Total Deaths")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width = FALSE,
    position = "left"
  ) %>%
  column_spec(1, width = "4cm", bold = TRUE) %>%
  column_spec(2, width = "4cm")                 

# Plotting the rate of cancer deaths over time
library(ggplot2)
ggplot(cancer_trend, aes(x = Year, y = Total_Deaths)) +
  geom_line(color = "aquamarine3", linewidth = 1.2) +
  geom_point(color = "black") +
  labs(title = "Trend of Cancer Deaths Over Time",
       x = "Year",
       y = "Total Cancer Deaths") +
  theme_minimal()
```

In this data we can see that the deaths caused by cancer keep increasing throughout the years. We can see that there is a dip in 2004, this is due to some states statistical data not meeting the requirements to be included to the US data. Since this happened, the amount of deaths decreased due to some states not meeting the requirements to input their data. 


# We know the adjusted death rate for the data, but what does the rate look like when we subtract "all causes'?

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)

all_causes <- data %>%
  filter(Cause.Name == "All causes") %>%
  select(State, Year, TotalDeaths = Deaths)

known_cause_totals <- data %>%
  filter(Cause.Name != "All causes") %>%
  group_by(State, Year) %>% 
  summarise(KnownCauseDeaths = sum(Deaths), .groups = "drop")

residual_deaths <- all_causes %>%
  left_join(known_cause_totals, by = c("State", "Year")) %>%
  mutate(UnclassifiedDeaths = TotalDeaths - KnownCauseDeaths)


residual_deaths %>%
  filter(State != 'United States') %>% 
  mutate(
    UnclassifiedDeaths = TotalDeaths - KnownCauseDeaths, 
    across(c(TotalDeaths, KnownCauseDeaths, UnclassifiedDeaths), 
           ~ format(., big.mark = ","))
  ) %>% 
  select(State, Year, TotalDeaths, KnownCauseDeaths, UnclassifiedDeaths) %>%
  head(10) %>% 
  kable(
    col.names = c("State", "Year", "Total Deaths", "Known Cause Deaths", "Unclassified Deaths"),
    align = c("l", "c", "r", "r", "r")
  ) %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width = FALSE,
    position = "left"
  ) %>%
  column_spec(1, width = "3cm", bold = TRUE) %>%
  column_spec(2, width = "2cm") %>%
  column_spec(3:5, width = "3cm")
      


# Plot 1: Deaths by known causes
plot1 <- data %>% 
  filter(State != "United States", Cause.Name != "All causes") %>% 
  ggplot(aes(x = Year, y = Deaths, fill = Cause.Name)) +
  geom_col(position = "stack") +
  labs(
    title = "Known Death Causes in U.S. States",
    x = "Years Since 2001",
    y = "Deaths",
    fill = "Cause of Death"
  ) +
  theme_minimal() +
  scale_x_discrete(guide = guide_axis(angle = 80)) +
  scale_fill_brewer(palette = "Paired") +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2))  

# Plot 2: Unexplained deaths
unclassified_data <- residual_deaths %>%
  filter(State != "United States") %>%
  mutate(Cause.Name = "Unclassified") %>%
  select(State, Year, Cause.Name, Deaths = UnclassifiedDeaths)

plot2 <- unclassified_data %>% 
  ggplot(aes(x = Year, y = Deaths)) +
  geom_col(position = "stack") +
  labs(
    title = "Unexplained Deaths",
    x = "Years Since 2001",
    y = "Deaths",
    fill = "State"
  ) +
  theme_minimal() +
  scale_x_discrete(guide = guide_axis(angle = 80)) +
  theme(legend.position = "none")   

combined <- (plot1 + plot2) +
  plot_layout(ncol = 2)

final_plot <- combined + plot_annotation(tag_levels = 'A')
final_plot
```
This plot represents the total amount of deaths through these diseases, but also the amount of unexplained deaths. The unexplained deaths can be due to the data being specifically based on the top 10 deaths in the United States. The total amount of deaths is from all residents death certificates that were filed through this time. 

## **What is age adjusted death rate?**
- Different states or regions have different age structures.

- Older populations naturally have higher death rates so comparing raw death rates across states would be misleading.

- To make fair comparisons across the states, public health stats use age adjustment instead of uing raw numbers. 



# What is the death rate by cause and state?

```{r echo = FALSE, warning=FALSE, message=FALSE}
library(plotly)
p <- data %>%   filter(State !="United States", Cause.Name != "All causes") %>% 
  group_by(State, Cause.Name) %>% summarise(RatebyState = mean(Age.adjusted.Death.Rate)) %>% 
  ggplot(aes(y = Cause.Name, x = RatebyState, color = Cause.Name, text = State)) + geom_jitter() +
  labs(title = "Death Rate by Cause and State",
       x = "Average Age-Adjusted Death Rate",
       y = "Cause of Death") +
  theme_minimal()
ggplotly(p)
```



```{r echo=FALSE}
library(dplyr)
library(ggplot2)

heart_disease_data <- data %>%
  filter(Cause.Name == "Heart disease")

ggplot(heart_disease_data, aes(x = as.factor(Year), y = reorder(State, -Age.adjusted.Death.Rate), fill = Age.adjusted.Death.Rate)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "plasma", name = "Death Rate") +
  labs(
    title = "Heatmap: Age-Adjusted Death Rate for Heart Disease",
    x = "Year",
    y = "State"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```


## How do the Amount of Deaths and Age Adjusted Death Rate Compare?

```{r echo = FALSE}
#COMPARE DEATH AND AGE AJUSTED

COMPARISON <- data %>%
  select(Year, Age.adjusted.Death.Rate, Deaths,Cause.Name,State) %>%
  filter(Cause.Name !='All causes', Cause.Name != 'Cancer', Cause.Name != 'Heart disease', State == 'United States') %>%
  ggplot(aes(x = Deaths, y = Age.adjusted.Death.Rate, color = Year)) + geom_jitter(alpha = .5) + facet_wrap(~Cause.Name) + coord_cartesian(xlim = c(0, 800000)) + 
  labs(
    x = 'Total Amount of Deaths',
    y = 'Age Adjusted Death Rate',
    title = 'Age Adjusted Death Rates Through All Diseases'
  ) + coord_cartesian(xlim = c(0,169000))

COMPARISON
```
When looking at all these graphs throughout the time, we can see the amount of deaths and the Age Adjusted Death Rate together. We see that through many of the death causes like CLRD, Stroke, Diabetes, Influeza and Pneumonia there has been a decrease throughout time. We see a huge decrease in the cause stroke also. Strokes can be caused if you have diabetes, it is an underlining of strokes because it can damage your blood vessels and cause your blood to cloth. We can see that diabetes has also decrease. Therefore these two are somewhat reflecting off of each other to an extent.  

## A Closer Look Into Heart Disease and Cancer

```{r echo = FALSE}


COMPARISON.2 <- data %>%
  select(Year, Age.adjusted.Death.Rate, Deaths,Cause.Name,State) %>%
  filter(Cause.Name == c('Cancer','Heart disease'), State == 'United States') %>%
  ggplot(aes(x = Deaths, y = Age.adjusted.Death.Rate, color = Year)) + geom_jitter(size = 2.5) + facet_wrap(~Cause.Name) + labs(
    x = 'Total Deaths',
    y = 'Age Adjusted Death Rate',
    title = 'Age Adjusted Death Rate Decrease'
  ) + theme(panel.spacing = unit(1, 'cm', data = NULL)) + scale_color_gradient(low = 'red', high = 'turquoise')

COMPARISON.2

```

When we look at this data, we can notice that there has been a change in Heart Disease and Cancer throughout these years. When I searched it up a bit more, it says that there can be changes throughout the years due to better technology. Due to better technology, these diseases are faster to detect. Therefore the age gap between the amount of deaths is closing in. Creating a greater percentage for those who have an older population. While deaths may be high, this can be due to the population increasing over the time. 

## A Closer Look Into Florida

```{r echo = FALSE}
FLORIDA_RATE <- data %>%
  select(Year, Age.adjusted.Death.Rate, Deaths,Cause.Name,State) %>%
  filter(Cause.Name == "Alzheimer's disease", State == 'Florida') %>%
  ggplot(aes(x = Deaths, y = Age.adjusted.Death.Rate, color = Year)) + geom_jitter() + facet_wrap(~Cause.Name) + coord_cartesian(xlim = c(0, 800000)) + 
  labs(
    x = 'Total Amount of Deaths',
    y = 'Age Adjusted Death Rate',
    title = "Age Adjusted Death Rates for Alzheimer's disease",
    subtitle = 'In Florida'
  ) + coord_cartesian(xlim = c(0,10000))

FLORIDA_RATE
```
While presenting we were asked about Florida specifically, mainly about the age in Florida. We were not given ages, but we can look at the Age Adjusted Death Rate and see that the percentages have increased. Alzheimer's disease is mainly diagnosed on people of older age. (Ages 65+) This disease has had an increase in Florida, so yes we can say that older people do live in Florida. Before though, not many people. The Age Adjusted Death Rate didn't see in increase until 2011, therefore now there are probably more older people moving to Florida. 

## LIMITATION

This project has been very interesting to work on. There are some limitations on it. I think it would've been better if our data also included ages. We think this would've helped a lot in some of the questions that we were receiving from people. Ages play a big roll into these diseases, we are given a bit by the Age Adjusted Death Rate but not to the fullest (Specifics on age range). I also think that it would've been nice to have the specific type of Cancer. We are given a more generalized name for Cancer, it is all in one category instead of being a general Cancer name and a specific Cancer name. 

## SUMMARY


